---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r, echo=FALSE}
library (datasets)
library (data.table)
#library(kernlab)
library (chron)
library (lubridate)
library (dplyr)
library (ggplot2)
library(data.table)
```

## Loading and preprocessing the data

The data does not include values for some days. (NA)

First read in the data and name the columns appropriately.

```{r}
stepdat <- read.table("./activity.csv",sep = ",",skip = 1, colClasses = c("numeric",
                                                                          "Date","numeric"))
cnames <- c("step","date","interval")
names(stepdat) <- make.names(cnames)
head (stepdat)
```

## What is mean total number of steps taken per day?
Determine the total for all 61 days and name the columns of the new dataframe "daytotal"
Eliminate days with NA values

```{r}
daytotal <- aggregate(list(stepdat$step), list(stepdat$date),sum, na.rm = TRUE)
head (daytotal)
names(daytotal) <- make.names(c("date","daily_total_steps"))
```

Make a histogram that shows a distribution of steps

```{r, echo=FALSE}
hist(daytotal$daily_total_steps, col="blue", main="Total Daily Steps", xlab="Daily Total Steps", 
     breaks = 25, xlim = c(0,25000))
```
Meadian and Mean values are observed in the table below:
```{r, echo=FALSE}
summary (daytotal)
```
Determine number of NAs in the dataset
```{r, echo=FALSE}
totalNA <- sum(is.na(stepdat$step))
print (totalNA)

print (totalNA)
```

Approach 1 for handling missing values:
For days that have all NAs replace with the mean value for all 61 days
Mean is 9354 steps as observed in summary

To do that: 
Step 1: Replace 0s in daytotal with the 9354 (daily mean of total steps)

```{r, echo=FALSE}
head (daytotal)
tmpdailytotalsteps <- data.frame(daytotal$daily_total_steps)
tmpdailytotalsteps[(tmpdailytotalsteps == 0)] <- round(mean(daytotal$daily_total_steps),0)
names(tmpdailytotalsteps) <- make.names("dailytotalsteps")
head (tmpdailytotalsteps)
```

Combines the columns to the daytotal dataframe
```{r, echo=FALSE}
daytotal <- cbind (daytotal, tmpdailytotalsteps)
head (daytotal)
```
Create a histogram that compares approach 1 (removing NAs) to approach 2 (substitiuting NA w/ avg values)
```{r, echo=FALSE}
par(mfrow = c (2,1))
hist(daytotal$dailytotalsteps, col="green", main="Total Daily Steps with modified NA values", 
     xlab="Daily Total Steps", breaks = 25, xlim = c(0,25000))
hist(daytotal$daily_total_steps, col="blue", main="Total Daily Steps with eliminating NA values", 
     xlab="Daily Total Steps", breaks = 25, xlim = c(0,25000))
summary (daytotal)
```
Notice the change in median and mean values from the 2 approaches.
```{r, echo=FALSE}
summary (daytotal)
```

## What is the average daily activity pattern?

Q2: Convert the intervals to factors so that we can see which interval has largest number of steps.
```{r, echo=FALSE}
stepdat <- mutate (stepdat, interval = factor(interval))
str (stepdat)
```
Above, we notice 288 levels

Calculate the mean for each interval across all days.
However since, few days have no data I will replace "NAs" for those days with the 
average of daily datasteps (9354) divided by the number of intervals (288)

```{r, echo=FALSE}
stepdat[is.na(stepdat)] <- round(mean(daytotal$dailytotalsteps)/288)
```


Question 2 - First using weekdays function, determine the days of the week for all 61 days.
```{r, echo=FALSE}
intervalavg <- data.frame (tapply(stepdat$step, stepdat$interval, mean))
names(intervalavg) <- make.names("interval_mean_for_all_days")
```

Create the timeline graph across intervals
```{r, echo=FALSE}
par(mfrow = c (2,1))
with (intervalavg, plot(interval_mean_for_all_days, 
                        main = "Mean Interval Steps", type = "l", 
                        xlab ="Interval", ylab = "Steps"))
```

## Are there differences in activity patterns between weekdays and weekends?
Quesstion 2 - Create factors from dayofweek variable, so that we can do analysis.
First identify which day of the week.
```{r, echo=FALSE}
daytotal <- mutate (daytotal, dayofweek = weekdays(date))
daytotal$weekend = is.weekend(daytotal$date)
str(daytotal)
```

Calculate the total steps for weekend or weekday and compare workkday (FALSE) and weekend (TRUE)
```{r, echo=FALSE}
stepsbydaytype  <- data.frame (tapply(daytotal$dailytotalsteps, daytotal$weekend, mean))
names(stepsbydaytype) <- make.names("meansteps")
print (stepsbydaytype)
par(mfrow = c (1,1))
with (daytotal, boxplot(dailytotalsteps ~ weekend,
                        main = "Mean Interval Steps by WorkDay and Weekend", 
                        xlab ="Type of Day", ylab = "Steps"))